Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2018-02-20 paper/manuscript.tex
fully incorporated figs with 5FYL, 5VN3, and min

\begin{figure}
\begin{fullwidth}
{\bf \Large A} \hspace{0.61\textwidth} {\bf \Large B} \\
\hspace{0.01\textwidth} 
\includegraphics[width=0.58\textwidth]{figures/prefs_distance/prefs_distance.pdf}
\hspace{0.03\textwidth}
\includegraphics[clip=true, trim=0in 0in 0in 1.2in, width=0.68\textwidth]{figures/distance_distribution.pdf}
\vspace{0.15in}
\\ 
{\bf \Large C} \\
\includegraphics[width=1.3\textwidth]{figures/shifted_sites.pdf}
\caption{\label{fig:prefsdist}
Env sites with shifted amino-acid preferences between BG505 and BF520.
Note that the preferences have been re-scaled using the stringency parameters in \TABLE{phydms} to enable direct comparison across Envs.
{\bf (A)} Calculation of the corrected distance between the amino-acid preferences of BG505 and BF520 at four example sites.
We have triplicate measurements for each Env.
We calculate the distance between each pair of replicate measurements, and group these into comparisons \emph{between} the two Envs and \emph{within} replicates for the same Env.
We compute the root-mean-square distance (RMSD) for both sets of comparisons, which we denote as RMSD$_{\rm{between}}$ and RMSD$_{\rm{within}}$.
The latter quantity is a measure of experimental noise.
The noise-corrected distance between Envs at a site, RMSD$_{\rm{corrected}}$, is simply the distance between the two Envs minus this noise.
{\bf (B)} The bottom distribution (orange) shows the corrected distances between BG505 and BF520 at all alignable sites.
The next distribution (blue) is a null generated by computing the corrected distances on all randomizations of the replicates among Envs.
The top two distributions (green) compare Env to the non-homologous influenza hemagglutinin (HA) protein~\citep{doud2016accurate} simply putting sites into correspondence based on sequence number.
We compute the $P$-value that a site has shifted between BG505 and BF520 as the fraction of the null distribution that exceeds that shift, and identify significant shifts at a false discovery rate (FDR) of 0.1 using the method of \citet{benjamini1995controlling}.
Using this approach, 30 of the 659 sites have significant shifts (corrected distance $\ge$0.22).
{\bf (C)} All sites that have significantly shifted their amino-acid preferences at an FDR of 0.01.
For each site, the logo stacks show the across-replicate average preferences for BG505 and BF520.
The wildtype amino acid for that Env is indicated using the small black letters above each logo plot; note how the wildtype amino acid is frequently but not always the most preferred one.
The sites are sorted by the magnitude of the shift.
}
\figdata{\label{figdata:prefsdist}The corrected distances between BG505 and BF520 at each site are in \texttt{BG505\_to\_BF520\_prefs\_dist.csv}.}
\end{fullwidth}
\end{figure}

@@

\begin{figure}
{\bf \Large A} \\
\includegraphics[width=0.8\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
\\
{\bf \Large B} \hspace{0.28\textwidth} {\bf \Large C} \hspace{0.28\textwidth} {\bf \Large D} \\
\includegraphics[width=0.28\textwidth]{figures/rsa_vs_shifts.pdf}
\hspace{0.01\textwidth}
\includegraphics[width=0.28\textwidth]{figures/shifts_proximity.pdf}
\hspace{0.01\textwidth}
\includegraphics[width=0.4\textwidth]{figures/conservation_vs_shifts.pdf} 
\caption{\label{fig:shiftfeatures}
Characteristics of significantly shifted sites.
{\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from white to red according to the magnitude of the mutational shift at each site (red indicates large shift).
Sites that are significantly shifted according to \FIG{prefsdist}B are in spheres, and all other sites are in cartoon representation.
{\bf (B)} 
There is no significant difference in the relative solvent accessibility of sites that have and have not undergone significant shifts.
The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
{\bf (C)} 
Sites of significant shifts are clustered in Env's structure.
Shown is the distance of each significantly shifted and not-shifted site to the closest other shifted site.
{\bf (D)}
Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
\FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
All panels only show the 577 sites that are resolved in the crystal structure.
Of the 30 significantly shifted sites in \FIG{prefsdist}B, only 28 are crystallographically resolved.
Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
$P$ values were calculated using the Mann-Whitney $U$ test.
}
\figsupp[\label{figsupp:subs_shifts}Statistical testing of whether significantly shifted sites are more likely to have substituted between BG505 and BF520.]
{The sites of significant shifts as determined in \FIG{prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
{\input{figures/shifts_vs_subs_table}}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0.7in, width=0.68\textwidth]{figures/entrenchment.pdf}}
\caption{\label{fig:entrenchment}
Entrenchment of substitutions during Env evolution.
There are 12,521 possible amino-acid mutations at the 659 mutagenized sites alignable between BG505 and BF520.
The blue densities show the effects of all these mutations to each Env.
The orange densities show the effects of just the 92 mutations that convert BG505 to BF520 or vice versa.
In the absence of entrenchment, mutating a site in BG505 to its identity in BF520 should have the opposite effect of mutating the site in BF520 to its identity in BG505.
In this case, we would expect the BF520$\rightarrow$BG505 distribution to be the mirror image of the BG505$\rightarrow$BF520 distribution -- and both distributions should be centered around zero if the two Envs are equivalently functional.
Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions are often entrenched in the Env in which they have fixed.
The effect of a mutation is quantified as the log of the ratio of the site's preference for the mutant amino acid to the preference for the wildtype amino acid.
}
\end{figure}

@@

}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}

>>>>

\begin{figure}
\begin{fullwidth}
{\bf \Large A} \hspace{0.61\textwidth} {\bf \Large B} \\
\hspace{0.01\textwidth} 
\includegraphics[width=0.58\textwidth]{figures/prefs_distance/prefs_distance.pdf}
\hspace{0.03\textwidth}
\includegraphics[clip=true, trim=0in 0in 0in 0.4in, width=0.65\textwidth]{figures/distance_distribution.pdf}
\vspace{0.15in}
\\ 
{\bf \Large C} \\
\includegraphics[width=1.3\textwidth]{figures/shifted_sites.pdf}
\caption{\label{fig:prefsdist}
Env sites with shifted amino-acid preferences between BG505 and BF520.
Note that the preferences have been re-scaled using the stringency parameters in \TABLE{phydms} to enable direct comparison across Envs.
{\bf (A)} Calculation of the corrected distance between the amino-acid preferences of BG505 and BF520 at four example sites.
We have triplicate measurements for each Env.
We calculate the distance between each pair of replicate measurements, and group these into comparisons \emph{between} the two Envs and \emph{within} replicates for the same Env.
We compute the root-mean-square distance (RMSD) for both sets of comparisons, which we denote as RMSD$_{\rm{between}}$ and RMSD$_{\rm{within}}$.
The latter quantity is a measure of experimental noise.
The noise-corrected distance between Envs at a site, RMSD$_{\rm{corrected}}$, is simply the distance between the two Envs minus this noise.
{\bf (B)} The bottom distribution (orange) shows the corrected distances between BG505 and BF520 at all alignable sites.
The next distribution (blue) is a null generated by computing the corrected distances on all randomizations of the replicates among Envs.
The top two distributions (green) compare Env to the non-homologous influenza hemagglutinin (HA) protein~\citep{doud2016accurate} simply putting sites into correspondence based on sequence number.
We compute the $P$-value that a site has shifted between BG505 and BF520 as the fraction of the null distribution that exceeds that shift, and identify significant shifts at a false discovery rate (FDR) of 0.1 using the method of \citet{benjamini1995controlling}.
Using this approach, 30 of the 659 sites have significant shifts (corrected distance $\ge$0.22).
{\bf (C)} All sites that have significantly shifted their amino-acid preferences at an FDR of 0.01.
For each site, the logo stacks show the across-replicate average preferences for BG505 and BF520.
The wildtype amino acid for that Env is indicated using the small black letters above each logo plot; note how the wildtype amino acid is frequently but not always the most preferred one.
The sites are sorted by the magnitude of the shift.
}
\figdata{\label{figdata:prefsdist}The corrected distances between BG505 and BF520 at each site are in \texttt{BG505\_to\_BF520\_prefs\_dist.csv}.}
\end{fullwidth}
\end{figure}

@@

\begin{figure}
{\bf \Large A} \hspace{0.6\textwidth} {\bf \Large B}  \\
\includegraphics[width=0.57\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
\hspace{0.06\textwidth}
\includegraphics[width=0.32\textwidth]{figures/rsa_vs_shifts.pdf}
\\
 {\bf \Large C} \hspace{0.45\textwidth} {\bf \Large D} \\
\includegraphics[width=0.45\textwidth]{figures/shifts_proximity.pdf}
\hspace{0.02\textwidth}
\includegraphics[width=0.53\textwidth]{figures/conservation_vs_shifts.pdf} 
\caption{\label{fig:shiftfeatures}
Characteristics of significantly shifted sites.
{\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from white to red according to the magnitude of the mutational shift at each site (red indicates large shift).
Sites that are significantly shifted according to \FIG{prefsdist}B are in spheres, and all other sites are in cartoon representation.
{\bf (B)} 
There is no significant difference in the relative solvent accessibility of sites that have and have not undergone significant shifts.
The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
{\bf (C)} 
Sites of significant shifts are clustered in Env's structure.
Shown is the distance of each significantly shifted and not-shifted site to the closest other shifted site.
{\bf (D)}
Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
\FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
All panels only show the 577 sites that are resolved in the crystal structure.
Of the 30 significantly shifted sites in \FIG{prefsdist}B, only 28 are crystallographically resolved.
Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
$P$ values were calculated using the Mann-Whitney $U$ test.
}
\figsupp[\label{figsupp:subs_shifts}Statistical testing of whether significantly shifted sites are more likely to have substituted between BG505 and BF520.]
{The sites of significant shifts as determined in \FIG{prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
{\input{figures/shifts_vs_subs_table}}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[clip=true, trim=0in 0in 0in 0in, width=0.68\textwidth]{figures/entrenchment.pdf}}
\caption{\label{fig:entrenchment}
Entrenchment of substitutions during Env evolution.
There are 12,521 possible amino-acid mutations at the 659 mutagenized sites alignable between BG505 and BF520.
The blue densities show the effects of all these mutations to each Env.
The orange densities show the effects of just the 92 mutations that convert BG505 to BF520 or vice versa.
In the absence of entrenchment, mutating a site in BG505 to its identity in BF520 should have the opposite effect of mutating the site in BF520 to its identity in BG505.
In this case, we would expect the BF520$\rightarrow$BG505 distribution to be the mirror image of the BG505$\rightarrow$BF520 distribution -- and both distributions should be centered around zero if the two Envs are equivalently functional.
Instead, mutating a site in either Env to its identity in the other Env tends to be deleterious, indicating that substitutions are often entrenched in the Env in which they have fixed.
The effect of a mutation is quantified as the log of the ratio of the site's preference for the mutant amino acid to the preference for the wildtype amino acid.
}
\end{figure}

@@

}{\includegraphics[width=\textwidth]{figures/selsites_rsa.pdf}}
